---
title: "more-dplyr-data-transformation"
output: html_document
date: '2023-10-18'
author: Sabrina Nardin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Code chunk

* PC: Cntr + Alt + I
* Mac: Cmd + Option + I

Pipe

* PC: Ctrl + Shift + M 
* Mac: Cmd + Shift + M

```{r}
library(tidyverse)
library(palmerpenguins)
```


```{r}
data("penguins")
```


## Explore the data

```{r}
# head and tail
head(penguins)
tail(penguins)
rbind(head(penguins, 5), tail(penguins, 5))
glimpse(penguins)
```

```{r}
# dataset summaries
str(penguins)
summary(penguins)
```


## Basic dplyr functions

select(): extracts columns/variables
```{r}
# select two specific variables using their names
select(penguins, species, body_mass_g)

# select all variables except two using the minus sign
select(penguins, -year, -island)

# select a range of contiguous/adjacent variables
select(penguins, species:bill_length_mm)

# select all variables except a range of contiguous ones
select(penguins, -(species:island))

# starts with
select(penguins, starts_with("bill"))

# ends with
select(penguins, ends_with("mm"))
```


filter(): extracts rows/observations based on columns/variables values
logical operators are commonly used with filter() 
```{r}
# ways to check variable values 
levels(penguins$species) 
unique(penguins$species)
count(penguins, species)

# filter by island and species
filter(penguins, (island == "Biscoe") & (species == "Gentoo"))
filter(penguins, island == "Biscoe", species == "Gentoo")

# filter by multiple conditions
filter(penguins, species %in% c("Adelie", "Gentoo"), island != "Biscoe")
```

filter() and missing data
```{r}
filter(penguins, is.na(sex))
filter(penguins, !is.na(sex))

# wrong code
# filter(penguins, sex == NA)
```

summarize() and missing data
```{r}
penguins %>%
  filter(species == "Adelie") %>%
  summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))
```


select() + filter()
```{r}
penguins %>% 
  select(species, sex, year, island) %>% 
  filter(island == "Torgersen")
```


relocate(): changes column order
```{r}
penguins %>%
  relocate(year, sex)

penguins %>% 
  relocate(year, .after = island)
```


arrange(): changes row order
```{r}
penguins %>%
  select(species, contains("bill")) %>%
  filter(!species == "Adelie") %>% 
  arrange(desc(bill_length_mm)) 

penguins %>%
  select(species, contains("bill")) %>%
  filter(!species == "Adelie") %>% 
  arrange(desc(species), bill_length_mm)
```


mutate(): changes values of columns and create new columns
```{r}
penguins %>% 
  mutate(bill_ratio = bill_length_mm / bill_depth_mm)
```


summarize(): create summary stats
```{r}
penguins %>% 
  summarize(body_mass_mean = mean(body_mass_g, na.rm = TRUE),
            body_mass_sd = sd(body_mass_g, na.rm = TRUE),
            n = n())

penguins %>% 
  filter(species == "Gentoo") %>% 
  summarize(body_mass_mean = mean(body_mass_g, na.rm = TRUE),
            body_mass_sd = sd(body_mass_g, na.rm = TRUE),
            n = n())
```


group_by(): aggregate categorical variables into groups, often used together with summarize()
```{r}
penguins %>% 
  group_by(island, sex) %>%
  drop_na(sex) %>%
  summarize(body_mass_mean = mean(body_mass_g, na.rm = TRUE))
```


count(): counts of observations by group
```{r}
# without count
penguins %>% 
  group_by(sex) %>% 
  summarize(n = n())

# with count
penguins %>% 
  count(sex)
```


## Other dplyr functions

across(): facilitates syntax for column-wise operations, use inside functions like summarize() or mutate()
```{r}
# number of distinct entries for columns without across
penguins %>%
  summarise(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_year = n_distinct(year))

# number of distinct entries for columns with across
penguins %>%
  summarise(across(c(species, island, year), 
                   n_distinct))

# the opposite
penguins %>%
  summarise(across(-c(species, island, year), 
                   n_distinct))

# everything
penguins %>%
  summarise(across(everything(), 
                   n_distinct))

# everything with sum of na
penguins %>%
  summarise(across(everything(), 
                   ~sum(is.na(.))
                   ))
```


crosstab of two variables
```{r}
# dplyr + tidyr
penguins %>%
  group_by(year, species) %>%
  summarize(n = n()) %>%  # number of obs. in a row
  spread(year, n)
```


## Renaming and recoding variables

rename(): changes names of variables with new_name = old_name syntax
```{r}
penguins %>% 
  rename(specie = species)

# ps: remember to save results in a new dataframe
```


mutate() + recode() with old_name = new_name syntax
```{r}
# check
count(penguins, sex)

# mutate + recode
penguins %>%
  mutate(sex = recode(
    sex,
    male = 1, 
    female = 0, 
    .default = NA_real_)) %>%
  count(sex)
```

mutate() + case_when()
```{r}
penguins %>%
  mutate(sex = case_when(
    sex == "female" ~ 0,  # old ~ new
    sex == "male" ~ 1
  )) %>%
  count(sex)
```
